---
title: "pcwOrd: (partial) (constrained) (weighted) ordination"
author: "PME"
date: "3/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Motivation

Many ordination packages exist for ecologists using R - `vegan` is excellent, for example. But none readily perform (to my knowledge) weighted, partial constrained ordinations.

The motivation for this is to analyze compositional ecological data - especially high throughput sequencing data - with methods that are robust, reproducible, and transparent. Compositional data contains only relative data (think relative abundance), and so requires a bit of extra care to analyze. The dominant approach is to perform a log-ratio transformation, then use principle components analysis - combined, this is called log-ratio analysis. For an enterance into this literature, see Greenacre and Aitchison (2002), Greenacre and Lewi (2009), and Quinn et al (2018). 

In particular, Greenacre and Lewi (2002) suggest using weighted log-ratio analysis, as unweighted log-ratio analysis is susceptible to noise in low-abundance features. Raw values of low-abundance features have higher relative variance than raw values of high-abundance features; therefore, the log-ratios of low abundance features are also noisy. When low-abundance features dominate a solution the distances and inferences made about samples is less robust to error and noise. Note that correspondence analysis also is suceptible to dominance by low-abundance features. 

Weighted log-ratio analysis down-weights these uncertain, low-abundance features. As a result, the observed distances among samples should be more reproducible across experiments.

`vegan` can perform constrained and partiald log-ratio analysis if the community matrix is log-transformed beforehand. However, it does not provide a straightforward way to weight columns and working with ordination objects is not intuitive. `easyCODA` allows weighted and constrained log-ratio analysis, but does not allow partialing of nuisance effects and does not provide hypothesis testing. `pcwOrd` allows partialed, constrained, and weighted ordinations, and also provides utility functions for investigating ordination objects, visualization, and hypothesis testing. The code is written to be self-documenting. Maybe you'll agree.

This isn't an R package yet, so load the functions with `source()`:

```{r message=FALSE}
# Libraries
libs = c('easyCODA',  # log-ratio analysis and CLR analysis
         'vegan',     # multivariate statistics
         'here')      # find files in your project directory

load_libs = function() {
  for(i in libs) {
    if (!require(i, character.only=TRUE)) {
      install.packages(i)
      library(i, character.only=TRUE)
    }
  }
}
load_libs()

# pcwOrd
source(here('Scripts', 'pcwOrd', 'pcwOrd.R'))

spider = readRDS(here('Scripts', 'pcwOrd', 'Spider.RDS'))
```


This vignette will demonstrate the features of pcwOrd: ordination, hypothesis-testing, and visualization. I'll use the `spiders` dataset from `mvabund`. This dataset has two tables:

1. A `data.frame` of environmental variables, to which I'll add a categorical version of `soil.dry`*. 
2. A `matrix` of community data. 

Certain pcwOrd functions use the rownames of the environmental and community data to cross-reference, so we want to ensure both tables have rownames. Usually these are something meaningful, like sample IDs.

*technically this should be ordinal, but we'll treat it as categorical.

```{r data}
# environmental data
envi = as.data.frame(spider$x)

# categorical soil wetness
cats = c('dry', 'damp', 'wet', 'soaking')
cc = ceiling(envi$soil.dry)
envi$soil.cat = factor(cats[cc],
                       levels=cats)

# community data
comm = as.matrix(spider$abund)

# rownames
nobs = nrow(comm)
row_names = c(letters[], LETTERS[])[1:nobs]
rownames(envi) = row_names
rownames(comm) = row_names

str(envi)
comm[1:8, 1:8]
```


# Basic Functions

## Ordination

We'll first perform a basic principle components analysis of the community with `pcwOrd()` and visualize the results with `plot_ord`:

```{r pca_pcword}
pca = pcwOrd(comm)
plot_ord(pca, 
         main='Spiders: Basic PCA')
```

The `pca` object is a list of class `pcwOrd` that contains a number of items. See the documentation for `pcwOrd`. Important ones are: `'Y_scaled'` - the community matrix after centering, scaling, and/or weighting but before further analysis; `'unconstrained'` - the singular value decomosition matrices (left `'u'` and right `'v'`) and values (`'d'`). All relevant information about this ordination can be calculated from these values.

### Eigenvalues

For example, if we want to calculate eigenvalues of this unconstrained ordination, we need to access the singular values (vector `d`) of the unconstrained solution:

```{r}
pca$unconstrained$d^2
```

### Axis Variances
We can view these same data by calling ord_variance, which summarizes the variance explained by each axis in the solution:
```{r}
ord_variance(pca)
```

### Scree plots
We can also visualize variances across axes in a scree plot with `ord_scree()`:

```{r}
ord_scree(pca)
```

### Calculate scores

If you want to make your own biplots, you can calculate row, column, centeroid, and biplot scores by calling `ord_scores`. Scores can be principle, standard, or contribution - with the latter being raw singular values (either `'u'` or `'v'`). To make plotting easy, we can also add grouping information from the `envi` table:

```{r}
grp = envi[, 'soil.cat', drop=FALSE] # preserve rownames
rowscores = ord_scores(pca, 
                       choice='row', 
                       scaling='principle', 
                       axes=c(1:3), 
                       add_grouping=grp)
head(rowscores)
```

When plotting both standard and principle scores, often you'll need to rescale one of them for visualization:

```{r}
colscores = ord_scores(pca, 
                       choice='column', 
                       scaling='standard', 
                       axes=c(1:3))
scaled_colscores = scale_scores(colscores, 
                                rowscores) # can also specify manual scaling factor

lapply(list(original = colscores, 
            rescaled = scaled_colscores), 
       head)
```

### Identify top features

Finally, it's useful to identify the top features. The default `scaling='contribution'` gives top contribution scores - i.e. the features that contribute most to the axes chosen. Here, we identify the spider taxa that have the five highest mean contributions across the first two axes:
```{r}
top_scores(pca, n=5, choice='column', scaling='contribution', axes=c(1:2))
```

# More ordinations

## Log-ratio analysis

To perform log-ratio analysis:

1. Remove zeros from your data. For simplicity, I'll replace zeros with a pseudocount of 0.5. Note that I am *not* adding the pseudocount as `x + 0.5`, as this will distort the ratios between observations.
2. Close the community matrix, so that all abundances are relative abundances and `rowSums(x) = 1`.
3. Perform the log-ratio transformation of your choice. The centered log-ratio (CLR), where values are centered on the geometric mean, is a common choice.
4. Run principle components analysis on the log-ratio transformed matrix
```{r}
# Pseudocounts
cc = comm
cc[cc==0] = 0.5 # replace zeros without adding to all values!

# Close the community
closed_comm = sweep(cc, 1, rowSums(comm), '/') # close to relative abundance

# perform centered log-ratio
geo_mean = apply(closed_comm, 1, 
                 function(x) {
                   exp(sum(log(x))/length(x))
                 })
lr_comm =  sweep(closed_comm, 1, geo_mean, '/')
lr_comm = log(lr_comm, 2)

lra = pcwOrd(lr_comm)
plot_ord(lra, 
         main='Spiders: Log Ratio Analysis')
```

## Weighted ordinations

To weight either rows or columns, for example by sequencing depth or column prevalence, you have three options:

1. Tell pcwOrd to calculate weights automatically by setting `weight_rows=TRUE` or `weight_columns=TRUE`. This uses row or column masses as weights (via `rowSums`, for example).
2. Specify a vector of length `nrow()` or `ncol()` with the weights.
3. Input the result of `easyCODA::CLR()`, which returns a weighted CLR transformation and row-weights. This approach is demonstrated below:

Here is a weighted LRA:

```{r}
wlr_comm = CLR(closed_comm) # CLR based on column-weighted geometric mean, plus column weights

w_lra = pcwOrd(wlr_comm)
plot_ord(w_lra, 
         main='Spiders: Column-Weighted LRA')
```

In this situation, the results of weighted and unweighted ordinations are approximately the same. See Greenacre and Lewi (2009) for a discussion of when weighted log-ratio analysis might be advantageous over a non-weighted log ratio analysis. 


## Constrained ordination

We might be interested in how these spiders associate with environmental variables. For example, we might be interested in whether moss and reflection structures spider communities. A constrained ordination like redundancy analysis will do this: pcwOrd will regress the community  matrix against moss and reflectivity, and then ordinate the fitted values. 

```{r}
X = envi[, c('moss', 'reflection')] # preserve rownames
rda = pcwOrd(comm, X=X)

plot_ord(rda, 
         main='Spiders by Moss Coverage', 
         color_legend_position=c(0.8, 0))
```

In this plot, colors show moss coverage, the first column in X. 

### Significance testing
To test whether spiders vary with wetness class, run a PERMANOVA on the ordination with `permute_ord()`:

```{r}
pp = permute_ord(rda)
pp
```

`permute_ord()` has a number of permutation models that parallel the options in `vegan::permutest()` - and `permute_ord()` will return the same results as `vegan::permutest()` with the same randomization seed. We see a p-value of `r pp$p_val` after `r pp$F_perm` permutations.

## Partial ordination
Say we want to look at how moass affects spider communities independently of moisture class. We can partial this out as Z:

```{r}
Z = envi[, 'soil.cat', drop=FALSE] # preserve rownames
X = envi[, c('moss', 'reflection'), drop=FALSE]

pcrda = pcwOrd(closed_comm, X, Z)

plot_ord(pcrda, 
         main='Spiders by moss, partialed by wetness', 
         shape_legend_position=c(-0.04, 0.01))
```

Each of the wetness categories is centered at zero, as expected due to partialing. Moss increases to the bottom left (darker red, and arrows). This model is highly significant:

```{r}
permute_ord(pcrda)
```

## Partialed, constrained, weighted ordination
The reason for this package is to perform partial, constrained, weighted ordinations. Here is a weighted log-ratio analysis of the spider community, constrained by moss and reflection, with moisture category partialed out:

```{r}
Y = CLR(closed_comm)

pcw_ord = pcwOrd(Y, X, Z)

plot_ord(pcw_ord, 
         main='Partialed, Constrained, Weighted Log-Ratio Analysis', 
         shape_legend_position=c(-0.4, 0.1))
```


# Comparison to `vegan` and `easyCODA`
Returning to eigenvalues: these eigenvalues are exactly the same as calculated by easyCODA's PCA, and solutions are the same:
```{r}
easy_pca = PCA(comm, weight=FALSE)
rbind(pcwOrd = pca$unconstrained$d^2,
      easyCODA = easy_pca$sv^2
)
```

Eigenvalues are not identical to `vegan` eigenvalues, because `pcwOrd` (and `easyCODA`) weights the initial community matrix rows by `1/sqrt(nrow)`, while `vegan` weights rows by `1/sqrt(nrow-1)`. You can access `vegan`-style weighting, which will agree with vegan results:

```{r}
veganlike_pca = pcwOrd(comm, as_vegan=TRUE)
vegan_pca = rda(comm)

rbind(pcwOrd = veganlike_pca$unconstrained$d^2, 
      vegan = vegan_pca$CA$eig)
```

These weightings don't change relative variance or relative distances, only absolute variances and distances. Generally, we're concerned with relative variances and distances, so the choice of (equal) weighting doesn't matter.

```{r}
list(
  coda_weightings = ord_variance(pca)$unconstrained,
  vegan_weightings = ord_variance(veganlike_pca)$unconstrained
)
```

`vegan::permutest()` gives the same as `pcwOrd::permute_ord()`, if you set the same seed.

```{r}
vegan_rda = rda(comm, X)
veganlike_rda = pcwOrd(comm, X, as_vegan=TRUE)

set.seed(445)
vegan_permutes = permutest(vegan_rda, 
                           permutations=999)
set.seed(445)
pcw_permutes = permute_ord(veganlike_rda)
keep = c('fitted', 'residuals', 'num_df', 'denom_df', 'F_stat', 'p_val')

list(vegan = vegan_permutes,
     pcwOrd = pcw_permutes[keep])
```

# Additional plotting options

## Added categories
Add groupings to any ordination. They can be continuous.
```{r}
row_group = envi[, 'soil.cat', drop=FALSE]
plot_ord(pca, 
         row_group=row_group, 
         main='PCA with groupings by soil moisture class')
```

## Different axes
Plot different axes
```{r}
plot_ord(pca, 
         row_group=row_group, 
         axes=c(3,5), 
         main='PCA axes 3 and 5',
         color_legend_position=c(0.6, 0))
```

## Limit column labels
Reduce clutter by only labeling the top columns (by plotted score)
```{r}
plot_ord(pca, 
         row_group=row_group, 
         max_labels=5)
```

## colors and shapes
Change the colors of text, groupings, arrows, and change shapes.
```{r}
new_colors = hcl.colors(4, palette='Viridis')

plot_ord(pca, 
         row_group=row_group, 
         max_labels=5, 
         discrete_scale=new_colors,
         col_text='green', 
         main='Fluorescent PCA')
```

If you constrain an ordination with categories, the plot function will automatically recognize these and plot them as named centeroids:

```{r}
X = envi[, 'soil.cat', drop=FALSE]
Y = CLR(closed_comm)

cat_cwLRA = pcwOrd(Y, X)

plot_ord(cat_cwLRA, 
         main='Spiders wLRA with Centeroids')
```

## Screeplots as grobs
Screeplots can be produced as grobs (via ggplot2) for downstream manipulation, saving as R objects, and arranging with `gridExtra`. If you want to do this with ordinations, you're on your own!

```{r}
ord_scree(cat_cwLRA, main='ggplot2 Scree Plot', as_grob=TRUE) + 
  theme_dark()
```

